```{r setup, echo=FALSE}
setwd("~/Documents/COVID-r-markdown/")
# knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library("data.table")
library(cdcfluview)
library(tinytex)
# tinytex::install_tinytex()
# tinytex::install_tinytex(force = TRUE)
library(tibble)
today <- as.character(today())
today <- as_tibble(today)
```

---
title: "Assessing the Impact of COVID-19 with the US Mortality Report"
output: html_document
---
<br>
Graphs last generated on `r today$value`
<br>
<br>
<br>
In seeking to quantify the impact of COVID-19 in the United States,  
**excess mortality** is thought by many to be one of the most useful metrics.
<br>
Assuming that COVID-19 is the cause for nearly all the extra deaths since the Spring of 2020,
the following graphs allow one to assess the relative impact of COVID-19 by looking at the 
increase in deaths from all causes since the onset of the pandemic. 
<br>
<br>
Let's first look at the current trend line and total deaths by year in a bar chart.
<br>
<br>
<br>
```{r setup, include=FALSE}
# Bring in CDC US mortality data
df <- read.csv('mortality.csv', header = TRUE, stringsAsFactors = FALSE)
df <- df[c('total_pni','all_deaths','week_start', 'week_end')]
df$week_end <- as.Date(df$week_end, format = "%Y-%m-%d")
df$week_start <- as.Date(df$week_start, format = "%Y-%m-%d")
df <- df %>% arrange(desc(week_end))

# Exclue the most recent rows as it seems to take the CDC several weeks to update 
# total death numbers
exc_date <- df$week_start[3]
exc_date
df <- subset(df, week_start < exc_date)

#remove any 2015 data - using week_start
exc_date2 <- as.Date("2016-01-01")
df <- subset(df, week_start > exc_date2)
```


```{r, echo=FALSE}
# plot US deaths due to any cause - trend line
p <- ggplot(df, aes(x=week_start, y=all_deaths)) + 
scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
geom_line(color="#69b3a2") + 
xlab("") + 
ylab("weekly deaths") +
theme(axis.text.x=element_text(angle=60, hjust=1)) +
ggtitle("Weekly US Deaths Due to Any Cause")
print(p)
```


```{r, include=FALSE}
# plot US deaths due to any cause - bar plot
df$year <- format(as.Date(df$week_start, format="%d/%m/%Y"),"%Y")
df_year <- df %>% 
group_by(year) %>% 
summarise(deaths = sum(all_deaths)) %>% 
as.data.frame()
df_year$deaths_label <- paste(round(df_year$deaths / 1000000,2), "M") 
df_year <- df_year[df_year$year != 2013,]
head(df_year)

y_min <- round(min(df_year$deaths)*0.75,0)
y_max <- round(max(df_year$deaths)*1.05,0)
```

```{r, echo=FALSE}
b <- ggplot(df_year, aes(x=year, y=deaths)) +
geom_bar(stat="identity", fill="steelblue") +
coord_cartesian(ylim = c(y_min, y_max)) + 
geom_text(aes(label=deaths_label), vjust=1.6, color="white",
position = position_dodge(0.9), size=3.0) +
ylab("year-to-date deaths") +
ggtitle("Yearly US Deaths Due to Any Cause")
print(b)
```
<br>
<br>
<br>
The total deaths by year bar plot doesn't give us a good feel for how the current year compares to previous years unless the graph was created right at year's end. Let's add another bar chart with year to date only deaths for each year.
<br>
<br>
```{r, include=FALSE}
# plot Year to Date US deaths due to any cause - bar plot
# get the current year and the number of weeks reported so far in that year
cyear <- year(exc_date)
cweek <- nrow(subset(df, format(week_end, "%Y") == cyear))

# filter the first cweek weeks in each year
df_ytd <- df %>%
  group_by(year) %>%
  slice_max(year, n=cweek) %>% 
  as.data.frame()

# get the first cweek rows by year
df_ytd <- data.table(df, key = "year")
df_ytd <- df_ytd[ , tail(.SD, cweek), by = year]

# sum by year first
df_ytd_sum <- df_ytd %>% 
  group_by(year) %>% 
  summarise(deaths = sum(all_deaths), n = n()) %>% 
  as.data.frame()

df_ytd_sum$deaths_label <- paste(round(df_ytd_sum$deaths / 1000000,2), "M") 
# df_ytd_sum <- df_ytd_sum[df_ytd_sum$year,]
df_ytd_sum

# calculate y axis limits
y_min <- round(min(df_ytd_sum$deaths)*0.75,0)
y_max <- round(max(df_ytd_sum$deaths)*1.05,0)
```

```{r, echo=FALSE}
b <- ggplot(df_ytd_sum, aes(x=year, y=deaths)) +
  geom_bar(stat="identity", fill="steelblue") +
  coord_cartesian(ylim = c(y_min, y_max)) + 
  geom_text(aes(label=deaths_label), vjust=1.6, color="white",
            position = position_dodge(0.9), size=3.0) +
  ggtitle("Year to Date US Deaths Due to Any Cause")
print(b)
```