---
title: "US Mortality Report"
output: html_document
---

In seeking to quantify the impact of COVID-19 in the United States, some have argued that 
**excess mortality** is the metric upon which most Americans can agree.

Assuming that COVID-19 is the cause for nearly all the extra deaths since the Spring of 2020,
the following graphs allow one to assess the relative impact of COVID-19 by looking at the 
increase in deaths from all causes since the onset of the pandemic. 
<br>
<br>

```{r setup, include=FALSE}
# knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
# tinytex::install_tinytex()
tinytex::install_tinytex(force = TRUE)

# Bring in CDC US mortality data
df <- read.csv('mortality.csv', header = TRUE, stringsAsFactors = FALSE)
df <- df[c('total_pni','all_deaths','week_start', 'week_end')]
df$week_end <- as.Date(df$week_end, format = "%Y-%m-%d")
df$week_start <- as.Date(df$week_start, format = "%Y-%m-%d")
df <- df %>% arrange(desc(week_end))

# Exclue the most recent rows as it seems to take the CDC several weeks to update 
# total death numbers
exc_date <- df$week_start[3]
exc_date
df <- subset(df, week_start < exc_date)

#remove any 2014 data - using week_start
exc_date2 <- as.Date("2015-01-01")
df <- subset(df, week_start > exc_date2)
```


```{r, echo=FALSE}
# plot US deaths due to any cause - trend line
p <- ggplot(df, aes(x=week_start, y=all_deaths)) + 
scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
geom_line(color="#69b3a2") + 
xlab("") + 
ylab("weekly deaths") +
theme(axis.text.x=element_text(angle=60, hjust=1)) +
ggtitle("Weekly US Deaths Due to Any Cause")
print(p)
```


```{r, include=FALSE}
# plot US deaths due to any cause - bar plot
df$year <- format(as.Date(df$week_start, format="%d/%m/%Y"),"%Y")
df_year <- df %>% 
group_by(year) %>% 
summarise(deaths = sum(all_deaths)) %>% 
as.data.frame()
df_year$deaths_label <- paste(round(df_year$deaths / 1000000,2), "M") 
df_year <- df_year[df_year$year != 2013,]
head(df_year)

y_min <- round(min(df_year$deaths)*0.75,0)
y_max <- round(max(df_year$deaths)*1.05,0)
```

```{r, echo=FALSE}
b <- ggplot(df_year, aes(x=year, y=deaths)) +
geom_bar(stat="identity", fill="steelblue") +
coord_cartesian(ylim = c(y_min, y_max)) + 
geom_text(aes(label=deaths_label), vjust=1.6, color="white",
position = position_dodge(0.9), size=3.0) +
ylab("year-to-date deaths") +
ggtitle("Yearly US Deaths Due to Any Cause")
print(b)
```